Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-11-26 rnni_convexity.tex
removed all the 'shortest caterpillar path' and similar terminology + todo to complete this fix

In this section we introduce an algorithm to compute paths between \emph{caterpillar trees}, which are trees where each internal node is adjacent to at least one leaf.
As a result caterpillar trees have only one \emph{cherry}, which is a pair of taxa that share their parent.
For introducing this algorithm we need the following definitions.
A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called a \emph{caterpillar path}.
If $p$ is the shortest path among all caterpillar paths between $T$ and $R$ we call it a \emph{shortest caterpillar path}, and denote the length of $p$ by $d_c(T,R)$, which we call the \emph{caterpillar distance}.

The special topology of caterpillar trees gives a natural preorder of taxa that uniquely represents a caterpillar tree.
That is, for a caterpillar tree $T = [\{x_1, x_2\}, \{x_1, x_2, x_3\}, \ldots, \{x_1, \ldots, x_n\}]$ the taxa are preordered according to the ranks of their parents such that $x_1 = x_2 \leq x_3 \leq \ldots \leq x_n$.
From now on when we say that a taxon is below another one we refer to this preorder.
Furthermore, we can use this preorder to create a list representation of caterpillar trees.
For the tree $T$ above this list is $[x_1, x_2, \ldots, x_n]$.
As we consider the taxon set $X = \{1, \ldots, n\}$, we assume that they are ordered in the list such that $x_1 < x_2$, according to their ordering as natural numbers.
This ensures that the list representation of caterpillar trees is unique.

%We are now going to explain what kind of $\rnni$ moves are possible between caterpillar trees and see how such a move changes the list representation of a tree.
%At first it is important to realise that the only possible $\rnni$ moves between caterpillar trees are $\nni$ moves, since there is no pair of internal nodes with rank difference one that is not connected by an edge.
%$\nni$ moves between caterpillar trees correspond to swaps of two neighboured elements in the list representation.
%The only exception of this are $\nni$ moves on the lowest internal edge, that is the edge connecting the internal nodes of rank one and two.
%There are two $\nni$ moves possible on this edge that result in caterpillar trees, while there is for every other edge only one $\nni$ move that results in a caterpillar tree.
%Corresponding to the $\nni$ moves on the lowest internal edge are exchanges of either the first or the second element with the third element in the list representation.

@@

\begin{lemma}
Let $T$ and $R$ be two caterpillar trees.
$T$ and $R$ have distance $d(T,R) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
\label{lemma:caterpillar_dist=diameter}
\end{lemma}

@@

For proving the other direction of the statement, let now $T$ and $R$ have caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:

@@

Induction step: $n \to n+1$\\
Induction hypothesis: For caterpillar trees $T, R$ on $n$ taxa with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ it is $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
Without loss of generality we can assume that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$.
Let $R$ be a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
It follows that in each loop of $\csort$ the maximum number of taxon swaps is reached which implies that every taxon will swap with taxon $n+1$ at some point.
We can conclude that this taxon $n+1$ is part of the cherry of $R$.
Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
$\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.

@@

\begin{proof}
We prove this Theorem by backwards induction on the caterpillar distance $l = d_c(T,R)$ between caterpillar trees $T$ and $R$.
For this backwards induction the maximum caterpillar distance $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:caterpillar_dist=diameter}) is the base case.

Induction basis: $l = \frac{(n-1)(n-2)}{2}$\\
With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, it is $d(T,R) = d_c(T,R)$.
Thus there is a shortest path between $T$ and $R$ that is a caterpillar path.

Induction step: $l+1 \to l$\\
Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that consists of caterpillar trees only, so the distance between these trees is $l+1$.

Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
We consider the shortest caterpillar path between these trees computed by $\csort$.
From now on we will consider the trees $T$ and $R$, as well as all other caterpillar trees within this proof, in their list representation $L_T$ and $L_R$ as introduced in Section~\ref{section:alg_csort}.

@@

So we can apply the induction hypothesis on $\hat T$ and $R$ and see that $d(\hat T,R) = l+1$.
If it now were $d(T,R) < l$, it would hold $d(\hat T,R) \leq d(T,R) + d(T,\hat T) < l + 1$ which contradicts the induction hypothesis.
We can conclude $d(T,R) = l$, which completes the proof for the case that $a$ is not one of the first two elements of $L_{T'}$.

Let us now consider the other case:
of the two taxa that are compared but do not swap positions, one is part of the cherry current tree.
Let $a$ and $b$ be the cherry taxa and $c$ their neighbour such that neither $a$ nor $b$ echanges with $c$.
Then the list looks like this:
\[[a, b, c, \ldots, x_k, \ldots, x_1]\]

If there is no swap of either $a$ or $b$ with $c$, we can construct a tree $\hat T$ with caterpillar distance $d_c(\hat T, R) = l+1$ and $d(T,\hat T) = 1$ in a similar way as in the previous case.
In the list representing the start tree $T$, either $a$ or $b$ must be neighbour of $c$.
This is due to the fact that we are currently considering the first comparison of elements where there is no swap.
We can assume without loss of generality that $a$ is neighbour of $c$ in $T$.
Then an exchange of $a$ and $c$ in $T$ yields a tree $\hat T$ with the desired properties.
As in the previous case we can then apply the induction hypothesis on $\hat T$ and conclude that there is a shortest path between $T$ and $R$ that only consists of caterpillar trees.
\end{proof}

>>>>

In this section we introduce an algorithm to compute paths between \emph{caterpillar trees}, which are trees where each internal node is adjacent to at least one leaf.
As a result caterpillar trees have only one \emph{cherry}, which is a pair of taxa that share their parent.
A path between two caterpillar trees that only consists of caterpillar trees is called a \emph{caterpillar path}.
We can identify a caterpillar tree $T = [\{x_1, x_2\}, \{x_1, x_2, x_3\}, \ldots, \{x_1, \ldots, x_n\}]$ with the list of its taxa $[x_1, x_2, \ldots, x_n]$, assuming that $x_1 < x_2$ as natural numbers (recall that the set of taxa is $\{1, \ldots, n\}$).

@@

\begin{lemma}
Let $T$ and $R$ be two caterpillar trees.
Then $d(T,R) = \frac{(n-1)(n-2)}{2}$ if and only if $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, where $d_c$ is the length of the shortest caterpillar path.
\label{lemma:caterpillar_dist=diameter}
\end{lemma}

@@

For proving the other direction of the statement, assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:

@@

Induction step:\\
Assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ implies $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
Without loss of generality we can assume that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$.
Let $R$ be a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
It follows that in each loop of $\csort$ the maximum number of taxon swaps is reached which implies that every taxon will swap with taxon $n+1$ at some point.
We can conclude that this taxon $n+1$ is part of the cherry of $R$.
Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
$\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
Hence, $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.

@@

\begin{proof}
We prove this theorem by backwards induction on the length of a shortest caterpillar path $d_c(T,R) = l$ ($l$ is the induction parameter) between trees $T$ and $R$.
It follows from the proof of Lemma~\ref{lemma:caterpillar_dist=diameter} that the maximum possible value for $l$ is $\frac{(n-1)(n-2)}{2}$, which forms the induction basis.
\todo{Lena: Make sure we ALWAYS use this terminology}
Lemma~\ref{lemma:caterpillar_dist=diameter} also implies in this case that $d(T, R) = d_c(T, R)$, hence a caterpillar path between $T$ and $R$ of length $d(T, R)$.

Let $T$ and $R$ be caterpillar trees such that $d_c(T, R) = l$.
We consider the shortest caterpillar path between these trees computed by $\csort$.
From now on we will consider the trees $T$ and $R$, as well as all other caterpillar trees within this proof, in their list representation $L_T$ and $L_R$ as introduced in Section~\ref{section:alg_csort}.

@@

The induction hypothesis for $\hat T$ and $R$ implies that $d(\hat T,R) = l+1$.
If it now were $d(T,R) < l$, it would hold $d(\hat T,R) \leq d(T,R) + d(T,\hat T) < l + 1$ which contradicts the induction hypothesis.
We can conclude that $d(T,R) = l$, which completes the proof for the case that $a$ is not one of the first two elements of $L_{T'}$.

Let us now consider the other case: of the two taxa that are compared but do not swap positions, one is part of the cherry current tree.
Let $a$ and $b$ be the cherry taxa and $c$ their neighbour such that neither $a$ nor $b$ exchanges with $c$.
Then the list looks like this:
\[[a, b, c, \ldots, x_k, \ldots, x_1]\]

If there is no swap of either $a$ or $b$ with $c$, we can construct a tree $\hat T$ with $d_c(\hat T, R) = l+1$ and $d(T,\hat T) = 1$ in a similar way as in the previous case.
In the list representing the start tree $T$, either $a$ or $b$ must be neighbour of $c$.
This is due to the fact that we are currently considering the first comparison of elements where there is no swap.
We can assume without loss of generality that $a$ is neighbour of $c$ in $T$.
Then an exchange of $a$ and $c$ in $T$ yields a tree $\hat T$ with the desired properties.
As in the previous case we can then apply the induction hypothesis on $\hat T$ and conclude that there is a shortest path between $T$ and $R$ that only consists of caterpillar trees.
\end{proof}
